---
title: "Temp_GZ"
author: "Mara Vukelić"
date: "2025-06-09"
output: html_document
---

#Doing multistate CMR analysis to see the effect of temperature on survival of West African giraffe population in the Giraffe Zone, Niger, 2005 - 2019
#Data for 2005-2018 are from 'ECMWF/ERA5/DAILY'




```{r}

rm(list = ls())

```


Loading libraries and encounter histories data
```{r}

library(tidyverse)
library(RMark)
library(here)


eh <- read.csv('./data/WA_giraffe_encounter_histories.csv', sep = ";", 
               colClasses = c("character", "character"))

```


Data processing and making design data list
```{r}

## processing the encounter history
giraffe_processed <- process.data(eh, model = "Multistrata", group = 'SEX', begin.time = 2005)

## create design data list 
giraffe_ddl <- make.design.data(giraffe_processed, parameters = list(S = list(pim.type = "time"), p = list(pim.type = "time"), Psi = list(pim.type = "constant")))

```


Fixing Psi (transition parameter) for parameters that can not transition from one to another:
```{r}


giraffe_ddl$Psi$fix <- NA

## juvenile to adult
giraffe_ddl$Psi$fix[giraffe_ddl$Psi$stratum == "1" & giraffe_ddl$Psi$tostratum == "3"] <- 0

## subadult to juvenile
giraffe_ddl$Psi$fix[giraffe_ddl$Psi$stratum == "2" & giraffe_ddl$Psi$tostratum == "1"] <- 0

## adult to juvenile
giraffe_ddl$Psi$fix[giraffe_ddl$Psi$stratum == "3" & giraffe_ddl$Psi$tostratum == "1"] <- 0

## adult to subadult
giraffe_ddl$Psi$fix[giraffe_ddl$Psi$stratum == "3" & giraffe_ddl$Psi$tostratum == "2"] <- 0


```



Testing the effect of number of days from March to NOvember, where the maximum daily temperature was above 35°C, on survival. Temperature data were extracted for the Giraffe Zone which was taken from O'Conner et al 2019 (Conclusion of the preliminary testing: Among various temperature stress metrics tested, survival was most consistently and negatively associated with the number of days with the maximum temperature above 35 °C outside the cool season (DJF)).

```{r}

library(dplyr)
library(ggplot2)
library(lubridate)

# Load the temperature data
TMaxDaily <- read.csv('./data/DailyMaxTemp_GZ_simple_05_18.csv')

# Ensure date column is in proper Date format
TMaxDaily <- TMaxDaily %>%
  mutate(date = as.Date(date),
         year = year(date),
         month = month(date))


```



```{r}


# Days with max temp ≥ 35°C, excluding Dec/Jan/Feb
hot_35_no_djf <- TMaxDaily %>%
  filter(max_temp_C >= 35, !month %in% c(12, 1, 2)) %>%
  count(year, name = "hot_35_no_djf")


# merge
giraffe_ddl$S <- giraffe_ddl$S %>%
  mutate(time = as.numeric(as.character(time))) %>%
  left_join(hot_35_no_djf, by = c("time" = "year")) %>%
  mutate(time = as.factor(time))  # restore to factor for modeling


# Calculate anomalies (z-scores)

### We are calling this new covariate N_hot_days !!!

hot_35_no_djf <- hot_35_no_djf %>%
  mutate(
    N_hot_days = scale(hot_35_no_djf)[, 1]
  )


# Merge with giraffe_ddl
giraffe_ddl$S <- giraffe_ddl$S %>%
  mutate(time = as.numeric(as.character(time))) %>%  # Ensure numeric time for joining
  left_join(hot_35_no_djf %>% select(year, N_hot_days),
            by = c("time" = "year")) %>%
  mutate(time = as.factor(time))  # Restore factor structure


```



```{r}


giraffe_analysis <- function(){
  
   S.1 <- list(formula = ~ SEX)
   
   S.2 <- list(formula = ~ time+SEX)
   S.3 <- list(formula = ~ N_hot_days+SEX)
   
   S.4 <- list(formula = ~ SEX+stratum)
   
   S.5 <- list(formula = ~ time+SEX+stratum)
   S.6 <- list(formula = ~ N_hot_days+SEX+stratum)
   
   S.7 <- list(formula = ~ SEX*stratum)
 
   S.8 <- list(formula = ~ SEX*stratum+time)
   S.9 <- list(formula = ~ SEX*stratum+N_hot_days)

   
   
   p.1 <- list(formula = ~ time+SEX)
   p.2 <- list(formula = ~ time*stratum)
   p.3 <- list(formula = ~ time+SEX+stratum)
   p.4 <- list(formula = ~ SEX+stratum*time)

  
  Psi.1 <- list(forumla = ~ -1 + stratum:tostratum)
  
  model.list <- create.model.list("Multistrata")
  
  mark.wrapper(model.list, data = giraffe_processed, ddl = giraffe_ddl, useddl = TRUE)
  
}

results_MSMR_N_hot_days_GZ_simple <- giraffe_analysis()

```


